---
title: "Lab_8"
format: pdf
editor: visual
---

```{r}
fna.data <- read.csv("~/Desktop/WisconsinCancer.csv")
wisc.df <- data.frame(fna.data, row.names = 1)
```

```{r}
# We can use -1 here to remove the first column
wisc.data <- wisc.df[,-1]
# Create diagnosis vector for later 
diagnosis <- as.factor(wisc.df$diagnosis)
```

Q1. How many observations are in this dataset?

There are 569 observations in this dataset

Q2. How many of the observations have a malignant diagnosis?

There are 212 observations that have a malignant diagnosis

Q3. How many variables/features in the data are suffixed with \_mean?

There are 10 variables which contain \_mean as a suffix.

```{r}
#Q1
nrow(wisc.df)
length(diagnosis)
#Q2
#wisc.df %>% filter(diagnosis == "M") %>% nrow()
table(wisc.df$diagnosis)
#Q3
length(grep("_mean", colnames(wisc.df), value = TRUE))
```

```{r}
# Check column means and standard deviations
colMeans(wisc.data)

apply(wisc.data,2,sd)
```

```{r}
wisc.hc <- hclust(dist(wisc.data))
plot(wisc.hc)
```

```{r}
# Perform PCA on wisc.data by completing the following code
wisc.pr <- prcomp(wisc.data, scale. = TRUE)
summary(wisc.pr)
biplot(wisc.pr)
plot(wisc.pr$x[, 1], wisc.pr$x[, 2], col = diagnosis, xlab = "PC1 (44.27%)", ylab = "PC2 (18.97%)")
#Scaling is important to normalize and compare varience in the data, so the results arent scued by variables with large variences in their data
sum_pca <- summary(wisc.pr)
barplot(sum_pca$importance[2,])
```

```{r}
library(factoextra)
fviz_eig(wisc.pr, addlabels = TRUE)
```

```{r}
pcs_res <- as.data.frame(wisc.pr$x)
ggplot(data = pcs_res, aes(x = PC1, y = PC2, color = diagnosis)) + geom_point() + theme_bw() + scale_color_manual(values = c("tan", "black"))
```

Q4. From your results, what proportion of the original variance is captured by thee first principle component?

The first component explains 44.27% of the variance in the dataset.

Q5. How many principle components (PCs) are required to describe at least 70% of the original variance in the data?

Three Pc's so (PC1, PC2, and PC3) are required to explain at least 70% of the variance in the original data.

Q6. How many principle components (PCs) are required to describe at least 90% of the original variance in the data?

Seven PC's (PC1, PC2, PC3, PC4, PC5, PC6, PC7) are required describe at least 90% of the variance of the data.

```{r}
biplot(wisc.pr)
```

Q7. What stands out to you about this plot? is it easy or difficult to understand? Why?

This plot is very messy and confusing. Due to the mess of sample and variable names it is really difficult to see or understand anything about what the data looks like, just too many inputs.

Q8. Generate a similar plot for principle components 1 and 3. What do you notice about these plots?

The points still segregate very well, however, it does seem that there is more overlap between the groups, as in the groups don't segregate very well.

```{r}
plot(wisc.pr$x[,1], wisc.pr$x[,3], col = diagnosis, xlab = "PC1", ylab = "PC3")

```

Q9. For the first principle component, what is the loading vector (ie wisc.pr\$rotation\[,1\]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC.

It is contributing 1.6% for the first principle component (Dimension 8).

```{r}
wisc.pr.hclust <- hclust(dist(wisc.pr$x[,1:3]), method = "ward.D2")
plot(wisc.pr.hclust)
```

Q.10 Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

I would say at a height of about 36 the clustering model has 4 clusters.

```{r}
plot(wisc.pr.hclust)
abline(h = 36, col = "red")
```

```{r}
plot(wisc.pr.hclust)
abline(h = 80, col = "red")
```

```{r}
grps <- cutree(wisc.pr.hclust, h = 80)
table <- grps
table(grps, diagnosis)
```

Q11. Can you find a better cluster vs diagnoses match by cutting into a different number of clusters between 2 and 10? How do you judge the quality of your result in each case?

The better cluster vs diagnosis match would be if you cut at two clusters. The quality of the result can be determined by evaluating the degree of overlap between the clustered groups and the diagnosis groups.

Q12. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

the ward.D2 method would be my preferred method, because it minimizes variance within clusters.

Q13. How well does the newly created model with two clusters separate out the two diagnoses?

The new model with the two clusters separates well and aligns nicely with the diagnoses results.

Q14. How well do the hierarchical clustering models you created in previous sections do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model with the vector containing the actual diagnoses.

It looks like the hierarchical clustering models do a fairly good job of separating the results with the appropriately selected K's.

Q15. Which of your analysis procedures resulted in a clustering model with the best specificity? How about sensitivity?

```{r}
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

```{r}
plot(wisc.pr$x[,1:2], col=diagnosis)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

Q16. Which of these new patients should we prioritize for follow up based on your results? Patient 2 should be followed up, more similar to malignant cancer patients.

```{r}
sessionInfo()
```
